KiCS2: A New Compiler from Curry to Haskell

### 2012-01-25 talks/ag2011/talk1.tex
Last years AG talks moved

\date{May 04, 2011}

@@

\begin{frame}[fragile]
\frametitle{Curry}
\begin{columns}[t]
\column{.45\textwidth}
\begin{itemize}
\item Lazy functional logic programing language.
\item Haskell-like syntax.
\item Extended by non-determinism, free variables, constraints, unification.
\end{itemize}
\begin{example}<2>
\begin{program}
> xorSelf aBool
False
False
\end{program}
\end{example}
\column{.45\textwidth}
\begin{block}{Running Code Example}
\begin{program}
data Bool = True | False

@@

\begin{frame}
\frametitle{KiCS2}
% Compilation strategies
%   - compile to logic language (Pakcs compiles to Prolog)
%   - compile to functional language (KiCS compiles to Haskell)
%   - compile to abstract machine (MCC compiles to C)
\begin{block}{Approach}
\begin{itemize}
\item Compile Curry programs into Haskell
\item Reuse features of Haskell (lazy evaluation, higher-order functions)
\item Benefit from mature Haskell compiler (GHC)
\end{itemize}
\end{block}
\pause
\begin{block}{Goals}
\begin{itemize}
\item Efficient execution of purely functional programs
\item Avoid unsafe operations to enable optimization (in contrast to KiCS)
\item Support for different search strategies (including parallel strategies)
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Representation of Non-Determinism}
\begin{itemize}
  \item No built-in non-determinism in Haskell.
  \item Idea: Explicit representation in data types as constructors.
\end{itemize}
\pause
\begin{block}{Curry}
\begin{program}
data Bool = True | False

@@

\begin{frame}[fragile]
\frametitle{Extension of Operations}
\begin{itemize}
  \item Operations have to deal with non-deterministic values.
  \item Non-determinism in arguments is propagated.
%   \item Purely functional Curry programs are Haskell programs (no runtime overhead) (first-order)
\end{itemize}
\pause
\begin{block}{Haskell}
\begin{program}
not True           = False
not False          = True
not (Choice x1 x2) = Choice (not x1) (not x2)

@@

\begin{frame}[fragile]
\frametitle{Example calls}
\begin{example}
\verb!not aBool!\\
$\leadsto$ \verb!not (Choice True False)!\\
$\leadsto$ \verb!Choice (not True) (not False)!\\
$\leadsto$ \verb!Choice False True!
\end{example}
\pause
\begin{example}
\verb!xorSelf aBool!\\
$\leadsto$ \verb!xor aBool aBool!\\
$\leadsto$ \verb!xor (Choice True False) aBool!\\
$\leadsto$ \verb!Choice (xor True aBool) (xor False aBool)!\\
$\leadsto$ \verb!Choice (not aBool) aBool!\\
$\leadsto$ \verb!Choice (Choice False True) (Choice True False)!
\end{example}
\begin{itemize}
  \item Yields two results \verb!True! and \verb!False!
  \item Only \verb!False! is correct w.r.t. the semantics of Curry.
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Higher-Order Functions (cont.)}
\begin{itemize}
\item Non-deterministic functions need an additional parameter
\item An identifier is passed to functional argument when applied
\end{itemize}
\begin{columns}[t]
\column{.45\textwidth}
\begin{block}{Curry}
\begin{program}
apply :: (a -> b) 
      -> a -> b
apply f x = f x

@@

\begin{columns}[t]
\column{.45\textwidth}
\begin{block}{Curry}
\begin{program}
> apply not True
False
\end{program}
\end{block}
\column{.45\textwidth}
\begin{block}{Haskell}
\begin{program}
> apply (\\x _ -> not x) True 1
False
\end{program}
\end{block}
\end{columns}
\vspace{3ex}
\begin{itemize}
\item Thus, higher-order functions are translated into
  \begin{itemize}
   \item a deterministic variant
   \item a non-deterministic variant
  \end{itemize}
\item Benchmarks show enormously improved performance
% \item Non-deterministic choice between functions is also possible.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Benchmark: Reverse of a List}
\begin{itemize}
\item Functional programs run with almost no overhead
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{reverse}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Benchmark: Reverse using foldr}
\begin{itemize}
\item Higher-order optimization dramatically improves performance
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{reverseho}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Benchmark: Non-Deterministic Permutation Sort}
\begin{itemize}
\item For non-deterministic programs, KiCS2 can compete with existing compilers
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{permsort}
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Benchmark: Halve a Peano Number}
\begin{itemize}
\item However, there is still much room for improvements
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{half}
\begin{program}
        half y | equal (add x x) y = x where x free
\end{program}
\end{center}
\end{frame}

@@

\begin{frame}
\frametitle{Conclusion}
\begin{itemize}
\item Transformation of a functional logic program into a functional program.
\item Resulting program can be optimized by the Haskell compiler.
\item Our Approach can compete with or outperform existing implementations.
\item A formal description of our approach can be found in our paper (WFLP 2011).
\end{itemize}
\end{frame}

>>>>

\date{June 08, 2011}

@@

\begin{frame}[fragile]
\frametitle{Curry}
\begin{columns}[t]
\column{.45\textwidth}
\begin{itemize}
\item Lazy functional logic programming language
\item Haskell-like syntax
\item Extended by non-determinism, free variables, constraints, unification
\end{itemize}
\begin{example}<2>
\begin{program}
> xorSelf aBool
False
False
\end{program}
\end{example}
\column{.45\textwidth}
\begin{block}{Running Code Example}
\begin{program}
data Bool = True | False

@@

\begin{frame}
\frametitle{KiCS2}
% Compilation strategies
%   - compile to logic language (Pakcs compiles to Prolog)
%   - compile to functional language (KiCS compiles to Haskell)
%   - compile to abstract machine (MCC compiles to C)
\begin{block}{Approach}
\begin{itemize}
\item Compile Curry programs into Haskell programs
\item Reuse features of Haskell (lazy evaluation, higher-order functions)
\item Benefit from mature Haskell compiler (GHC)
\end{itemize}
\end{block}
\pause
\begin{block}{Goals}
\begin{itemize}
\item Efficient execution of purely functional programs
\item Avoid unsafe operations to enable optimization (in contrast to KiCS)
\item Support for different search strategies (including parallel strategies)
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Representation of Non-Determinism}
\begin{itemize}
  \item No built-in non-determinism in Haskell
  \item Idea: Explicit representation in data types as constructors
\end{itemize}
\pause
\begin{block}{Curry}
\begin{program}
data Bool = True | False

@@

\begin{frame}[fragile]
\frametitle{Extension of Operations}
\begin{itemize}
  \item Operations have to deal with non-deterministic values
  \item Non-determinism in arguments is propagated
%   \item Purely functional Curry programs are Haskell programs (no runtime overhead) (first-order)
\end{itemize}
\pause
\begin{block}{Haskell}
\begin{program}
not True           = False
not False          = True
not (Choice x1 x2) = Choice (not x1) (not x2)

@@

\begin{frame}[fragile]
\frametitle{Example calls}
\begin{example}
\verb!not aBool!\\
$\leadsto$ \verb!not (Choice True False)!\\
$\leadsto$ \verb!Choice (not True) (not False)!\\
$\leadsto$ \verb!Choice False True!
\end{example}
\pause
\begin{example}
\verb!xorSelf aBool!\\
$\leadsto$ \verb!xor aBool aBool!\\
$\leadsto$ \verb!xor (Choice True False) aBool!\\
$\leadsto$ \verb!Choice (xor True aBool) (xor False aBool)!\\
$\leadsto$ \verb!Choice (not aBool) aBool!\\
$\leadsto$ \verb!Choice (Choice False True) (Choice True False)!
\end{example}
\begin{itemize}
  \item Yields two results \verb!True! and \verb!False!
  \item Only \verb!False! is correct w.r.t. the semantics of Curry
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Higher-Order Functions (cont.)}
\begin{itemize}
\item Non-deterministic functions need an additional parameter for fresh \verb!ID!s
\item The identifier is passed to the functional argument when applied
\end{itemize}
\begin{columns}[t]
\column{.45\textwidth}
\begin{block}{Curry}
\begin{program}
apply :: (a -> b) 
      -> a -> b
apply f x = f x

@@

\begin{columns}[t]
\column{.45\textwidth}
\begin{block}{Curry}
\begin{program}
> apply not True
False
\end{program}
\end{block}
\column{.45\textwidth}
\begin{block}{Haskell}
\begin{program}
> apply (\\x _ -> not x) True 1
False
\end{program}
\end{block}
\end{columns}
\vspace{3ex}
\begin{itemize}
\item Thus, higher-order functions are translated into
  \begin{itemize}
   \item a deterministic variant
   \item a non-deterministic variant
  \end{itemize}
\item Deterministic variant is used inside other deterministic functions
\item Benchmarks show enormously improved performance
% \item Non-deterministic choice between functions is also possible.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Benchmark: Reverse of a List}
\begin{itemize}
\item Functional programs run with almost no overhead
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{gfx/reverse}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Benchmark: Reverse using foldr}
\begin{itemize}
\item Higher-order optimization dramatically improves performance
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{gfx/reverseho}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Benchmark: Non-Deterministic Permutation Sort}
\begin{itemize}
\item For non-deterministic programs, KiCS2 can compete with existing compilers
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{gfx/permsort}
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Benchmark: Halve a Peano Number}
\begin{itemize}
\item However, there is still much room for improvements
\end{itemize}
\begin{center}
\includegraphics[width=9cm]{gfx/half}
\begin{program}
        half y | equal (add x x) y = x where x free
\end{program}
\end{center}
\end{frame}

@@

\begin{frame}
\frametitle{Conclusion}
\begin{itemize}
\item Transformation of a functional logic program into a functional program
\item Resulting program can be optimized by the Haskell compiler
\item Our approach can compete with or outperform existing implementations
\item A formal description of our approach can be found in our paper
\end{itemize}
\end{frame}
